A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-08 tex/codatadp.tex
Cleanup, more detail in proof.

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}

@@

%Given a set of disjoint paths $P$ where each $p \in P$
%we write $\interp{P}$ to denote the

\begin{lemma}[Solution Soundness]
If there exists a derivation tree with root node $\Ec$ containing a saturated node $\Fc$, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
%First, we construct a valuation $\Val$ as follows.
%Let $C$ be the set of variables occurring in the same equivalence class as a constructor term,
%and let $X$ be the set of variables occurring in $\mathcal{T}(\Fc)$ but not in $C$.

We assume that:
\begin{enumerate}
\item Each equivalence class of $\Fc$ contains at least one variable,
\item Each equivalence class $\ec{x}$ of $\Fc$ where $x : \tau$, $\tau \in \Nondata$ contains a constant,
\item Each equivalence class $\ec{x}$ of $\Fc$ where $x : \tau$, $\tau \in \Data \cup \Codata$ contains a constructor term.
\end{enumerate}
The first statement comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability.
The second statement can be assumed for $x : \tau$, since we know that $\tau$ is infinite.
The third statement can be assumed by \rem{TODO}.

We construct a model $\M$ of $\thD$ that interprets constructors and correctly-applied selectors in the obvious way,
interprets wrongly-applied selectors based on information in $\Fc$,
interprets the variables and $x$ of $\Fc$ based on a valuation $\Val$ which construct as follows.
For each equivalence class $e : \tau$, where $\tau \in \Nondata$ containing constant $c$,
$\Val(x) = c$ for each $x \in e$.
For each $e : \tau$, where $\tau \in \Data$ containing the constructor term $C( \vec u )$ where the equivalence classes $\ec{\vec u}$ contains the variables $\vec y$,
$\Val(x) = C( \Val( \vec y ) )$ for each $x \in e$.
Since \rn{Cyclic} does not apply, this definition is well-defined, and since \rn{Cong} does not apply, we have that $\Val{x} = \Val{y}$ if and only if $\ec{x} = \ec{y}$.
For each $e : \tau$, where $\tau \in \Codata$,
\rem{TODO}

We show that all equalities and disequalities in $\Fc$ are satisfied by $\M$.
Say $t_1 \teq t_2 \in \Fc$ where $t_1$ and $t_2$ have type $\tau$ where $\tau \not\in \Codata$,
and assume that $\ec{t_1}$ contains the variable $x$.
We show that $\M( t ) = \Val( x )$ for each term $t \in \ec{x}$.
In the case where $t$ is a variable, then $\M( t ) = \Val( x )$ since $t$ and $x$ are assigned identical values by construction.
In the case where $t$ is a constructor term $C( \vec u )$, say our construction built a valuation for $x$ based on the constructor term $C'( \vec u' ) \in \ec{x}$.
Since \rn{Unify$_2$} does not apply to $\Fc$, we have that $C = C'$.
Since \rn{Unify$_1$} does not apply to $\Fc$, we have that $\ec{\vec u} = \ec{\vec u'}$.
By structural induction, we have $\M( \vec u ) = \M( \vec u' )$, and thus $\M( t ) = \M( C( \vec u ) ) = \M( C'( \vec u ' ) ) = \Val( x )$.
In the case where $t$ is a selector term $s^j_k( u )$,
then since \rn{Split} does not apply, 
$\ec{u}$ must contain a term of the form $C_i( s^1_i( u ), \ldots, s^{a_i}_i( u ) )$ for some $i$.
Let $x_1, \ldots, x_n$ be the variables in equivalence classes $\ec{s^1_i( u )}, \ldots, \ec{s^{a_i}_i( u )}$.
By our construction, $\M( u ) = C_i( \Val( x_1 ), \ldots, \Val( x_{a_i} ) )$.
If $i = k$, then $\M( t ) = \M( s^j_k( u ) ) = \Val( x_j )$, which is $\Val( x )$ since $x_j \in \ec{s^j_i( u )} = \ec{x}$.
If $i \neq k$, then $\thD$ does not constraint the value $\M( t )$, and thus we interpret $s^j_k( \M( u ) )$ as $\Val( x )$.

%If $t$ and $s$ are both variables, then this follows since $\ec{t} = \ec{s}$.

Let $t \tneq s \in \Fc$.
Since \rn{Conflict} does not apply, we know $t \teq s \not\in \Fc$ and thus $\ec{t}$ and $\ec{s}$ are distinct.
Since by above, $\M( t ) = \M( s )$ if and only if $\ec{t}=\ec{s}$, we have $\M( t ) \neq \M( s )$, and thus $\M \models t \tneq s$.

By these two, 
%$\M \models_{\thD} \Fc$, and since $\Fc$ is a superset of $\Ec$, we have that $\Ec$ is satisfied by $\M$.
\qed
\end{proof}

>>>>

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}

@@

Given a set of disjoint paths $P$ where each $p \in P$
we write $\interp{P}$ to denote the value

\begin{lemma}[Solution Soundness]
If there exists a derivation tree with root node $\Ec$ containing a saturated node $\Fc$, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
We assume that:
\begin{enumerate}
\item Each equivalence class of $\Fc$ contains at least one variable,
\item Each equivalence class $\ec{x}$ of $\Fc$ where $x : \tau$, $\tau \in \Nondata$ contains a constant,
\item Each equivalence class $\ec{x}$ of $\Fc$ where $x : \tau$, $\tau \in \Data \cup \Codata$ contains a constructor term.
\end{enumerate}
The first statement comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability.
The second statement can be assumed for $x : \tau$, since we know that $\tau$ is infinite.
The third statement can be assumed by \rem{TODO}.

We construct a model $\M$ of $\thD$ that interprets constructors and correctly-applied selectors in the obvious way,
interprets wrongly-applied selectors based on information in $\Fc$,
interprets the variables and $x$ of $\Fc$ based on a valuation $\Val$, which is constructed as follows.
For each equivalence class $e : \tau$, where $\tau \in \Nondata$ containing constant $c$,
$\Val(x) = c$ for each $x \in e$.
For each $e : \tau$, where $\tau \in \Data$ containing the constructor term $C( \vec u )$ where the equivalence classes $\ec{\vec u}$ contains the variables $\vec y$,
$\Val(x) = C( \Val( \vec y ) )$ for each $x \in e$.
Since \rn{Cyclic} does not apply, this definition is well-defined. 
For each $e : \tau$, where $\tau \in \Codata$, 
$\Val( x ) = \interp{ \tpath{\Fc}{e} }$ for each $x \in e$.

Since \rn{Cong} does not apply, we have that:
\begin{eqnarray} \label{eqn:m-bij-dt}
\Val(x) = \Val(y) \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}

We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.

If $\tau \in \Nondata$, then $\M( t_1 ) = \M( t_2 )$ by construction.

If $\tau \in \Data$, then assume the equivalence class of $t_1$ and $t_2$ contains variable $x$.
We show that $\M( t ) = \Val( x )$ for each term $t \in \ec{x}$, which implies $\M \models t_1 \teq t_2$.
In the case where $t$ is a variable, then $\M( t ) = \Val( x )$ since $t$ and $x$ are assigned identical values by construction.
In the case where $t$ is a constructor term $C( \vec u )$, say our construction built a valuation for $x$ based on the constructor term $C'( \vec u' ) \in \ec{x}$.
Since \rn{Unify$_2$} does not apply to $\Fc$, we have that $C = C'$.
Since \rn{Unify$_1$} does not apply to $\Fc$, we have that $\ec{\vec u} = \ec{\vec u'}$.
By structural induction, we have $\M( \vec u ) = \M( \vec u' )$, and thus $\M( t ) = \M( C( \vec u ) ) = \M( C'( \vec u ' ) ) = \Val( x )$.
In the case where $t$ is a selector term $s^j_k( u )$,
then since \rn{Split} does not apply, 
$\ec{u}$ must contain a term of the form $C_i( s^1_i( u ), \ldots, s^{a_i}_i( u ) )$ for some $i$.
Let $x_1, \ldots, x_n$ be the variables in equivalence classes $\ec{s^1_i( u )}, \ldots, \ec{s^{a_i}_i( u )}$.
By our construction, $\M( u ) = C_i( \Val( x_1 ), \ldots, \Val( x_{a_i} ) )$.
If $i = k$, then $\M( t ) = \M( s^j_k( u ) ) = s^j_k( \M( u ) ) = \Val( x_j )$, which is $\Val( x )$ since $x_j \in \ec{s^j_i( u )} = \ec{x}$.
If $i \neq k$, then $\thD$ does not constraint the value of $\M( t )$ and since $s^j_k( \M( u ) )$ is otherwise unconstrained in our model construction since \rn{Cong} does not apply, 
we interpret $s^j_k( \M( u ) )$ as $\Val( x )$.

If $\tau \in \Codata$, \rem{TODO}

We show that all disequalities in $\Fc$ are satisfied by $\M$.
Say $t_1 \tneq t_2 \in \Fc$.
Since \rn{Conflict} does not apply, we know $t_1 \teq t_2 \not\in \Fc$ and thus $\ec{t_1}$ and $\ec{t_2}$ are distinct.
By the previous paragraph and statement~\ref{eqn:m-bij-dt}, $\M( t_1 ) \neq \M( t_2 )$, and thus $\M \models t_1 \tneq t_2$.

Thus, $\M \models \Fc$
and since $\Fc$ is a superset of $\Ec$, we have that $\Ec$ is satisfied by $\M$.
\qed
\end{proof}
